---
title: "EpiGPlot"
author: "Kevin Li"
date: "`r format(Sys.time(), '%d %b %Y')`"
output: 
 rmarkdown::html_vignette:
    toc: true
    number_sections: false
vignette: >
  %\VignetteIndexEntry{EpiGPlot}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
library(knitr)
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  comment = "#>"
)
set.seed(1)
```

**EpiGPlot** is a simple R package for producing a interpretative plot for epigenetic factors and their expression values against all other test sample genes. For plotting purposes and graphing linear regression models: ***plotEpigeneticEV***

For customizing plotting layouts and specifications: ***layoutEpigeneticEV***

For parsing csv or loading rda data for a certain gene and their expression values amongst other genes: ***parseEpigeneticData***, ***loadEpigeneticData***

Refer to `help(package = "EpiGPlot")` for further details.

To download and install the package use the following commands:

```{r setup}
require("devtools")
install_github("kevbobli224/EpiGPlot", build_vignettes = TRUE)
library("EpiGPlot")
```

To list all available exported functions for users, use:

``` {.r}
lsf.str("package:EpiGPlot")
```

<br>

## Details

Given a epigenetic factor, proteins with DNA/Histone/Chromatin remodeling; a list of expression values are obtained as well as their quantile against all genes. The referenced sample category can be plotted and visualized using linear regression methods.

## `plotEpigeneticEV` and `layoutEpigeneticEV`

In `plotEpigeneticEV`, the layout data returned from `layoutEpigeneticEV` is used to carry out visualization methods. `layoutEpigeneticEV` is required for the second round of parsing of the data from .csv/.rda because it contains unorganized data columns which the return of the layout function will serve as the layout data for various plotting functions from ggplot2 package.

User can specify various parameters from both `plotEpigeneticEV` and `layoutEpigeneticEV` in order to achieve an accurate plotting with specific parameters that are important for data interpretation.

### `layoutEpigeneticEv` 

To obtain a visualizable and plot-able data set, it must first be obtained from this function; where user can specify the operations and parameters that should the data be manipulated in a way that restricts future visual clutters. Furthermore, a simple centering algorithm can be performed if user wishes to center all expression values through the built-in base R function `scale`. This would allow data sets with extreme values to be normalized; this rescaling method will alter the range of expression values from 0 to 1.

Given a list of expression values of a epigenetic factor, it can be rescaled from 0 to 1 for the purpose of ease of plotting by

\begin{equation}

f(x)=\frac{x'-min(x)}{max(x)-min(x)}

\end{equation}

where $x$ is a vector of expression values and $x'$ is individual elements of $x$.
